Combinatorial interpretations of holonomic recurrences

### 2021-09-08 article/main.tex
Fix lambda

We also try to find~a combinatorial interpretation of the holonomic recurrence for trees of lambda terms in the de Bruijn notation.

@@

Every recursive structure can be described using~a generating function, which can be then transformed into~a holonomic recurrence. It allows us to quickly count objects of given structure with given size. At first glance, the recurrence seems to be closely related to the structure itself. It just seems like it should describe simple transformations resulting in constructing bigger objects from slightly smaller ones. In case of trees it could be for example adding vertices. However, there is no known method of interpreting it in the language of the combinatorial structures.

@@

Throughout the thesis we try to accumulate known results and techniques that may help solving the problem of finding combinatorial interpretations. We also propose~an approach that helped us to find~an intermediate solution for the problem of finding combinatorial interpretation for holonomic recurrence of lambda trees in the unary de Bruijn notation. We use the de Bruijn indices in order to get rid of letter-based variable names. Unary notation allows us to connect size of the tree to the values of the de Bruijn indices.

@@

Finally, in Section~\ref{sec:further_work}, we present our view on what can be done with the problem in the future. We believe that one day the problem will be fully solved and~a more generic approach to interpreting holonomic recurrences will emerge.

@@

\(\poly\) is~a set of all polynomials.

@@

\(V(T)\) is implicitly defined as~a set of nodes of the tree \(T\).

\(E(T)\) is implicitly defined as~a set of edges of the tree \(T\).

@@

\(e\) is implicitly assumed to be~an edge.

\(\troot(T)\) is implicitly defined as~a root of tree \(T\).

\(\tleft(v)\) is implicitly defined as~a left child of node \(v\).

\(\tright(v)\) is implicitly defined as~a right child of node \(v\).

\(\tSub(v)\) is implicitly defined as~a whole subtree rooted with node \(v\).

\(\tLeft(v)\) is implicitly defined as~a left subtree of node \(v\).

\(\tRight(v)\) is implicitly defined as~a right subtree of node \(v\).

@@

Uncle of node \(v\) is defined as~a sibling of the parent of \(v\).

We say that node \(v\) is~an oneling \(\textiff\) \(v\) has no siblings. Note that the root is oneling even if it has no parent.

@@

Leaf is~a node which has no children.

\(\n\) is implicitly defined as~a default identifier of~a node.

\(\no\) is implicitly defined as~an identifier of~a non-existing node (e.g.~if \(v\) is~a leaf, \(\tleft(v) = \n\)).

@@

Virtual node is~an artificially defined node that does not affect the structure definition (e.g.~we can define virtual leafs that are children of actual leaves of the binary tree. Such virtual leaves are not included when counting the number of children of the node, so the actual leaves are still called leaves).

Actual node is~a node that is not virtual.

We say that node \(v\) is~an virtual-oneling \(\textiff\) \(v\) has no siblings being virtual nodes.

@@

For~a binary tree, we say that node \(v\) is right (left) \(\textiff\) \(v\) is~a right (left) child of its parent.

For~a binary tree, we say that node \(v\) has~a right (left) parent \(u\) \(\textiff\) \(v\) is~a left (right) child of \(u\).

\(\pointed{v}\) is implicitly defined as~a pointed node \(v\). We can treat it as~a node with~an additional label \(\pointed{}\).

\(v\weighted{n}\) is implicitly defined as~a weighted node \(v\) with weight \(n\).

\(x_i\) is implicitly assumed to be~an \(i\)-th element of the sequence \(X\), where \(x\) can be any lower-case letter and \(X\) is upper-case \(x\).

Calligraphic \(\gf{X}\) is implicitly assumed to be~a generating function of the sequence \(X\) where \(X\) is any upper-case letter.

@@

We will be analyzing structures that can be expressed using recursive formulas. For example we can define~a full binary tree recursively as~a node having no children (i.e.~being~a leaf) or having exactly two subtrees being full binary trees. To express the disjoint alternative in the construction rule, we use \(+\). The reason will become obvious soon.

@@

\begin{definition}
The ordinary generating function of sequence \(A = (a_i)_{i=0}^{\infty}\) is defined as~a formal power series:
\[\gf{A}(z) = \sum_{i=0}^{\infty} a_i z^i .\]
\end{definition}

Note that in order to compute the \(k\)-th element of the sequence, namely \(a_k\), we just need to find~a constant corresponding to indeterminate \(z^k\).

If we have sequence counting some objects and corresponding generating function, we say that this function counts the objects.

\begin{example}
    \label{ex:bin_gf}
    Let's consider recursive definition of binary trees shown in Figure~\ref{fig:binary_recursion}. We can create~a functional equation to count trees with exactly n internal nodes. To do so, we make internal nodes contribute to the generating function with \(z^1 = z\) and leaves with \(z^0 = 1\):

@@

In order to get some intuition on how it works underneath, let's take~a look at how we can find the number of full binary trees of size \(2\). We can expand the recurrence~a bit:

@@

To visualize even more, let's take~a look at the line marked with \((*)\). The highlighted terms describe following tree:

@@

\begin{definition}[\cite{holotoolkit}]
    \label{def:holo_1}
    An infinite sequence \(A = (a_i)_{i=0}^{\infty}\) is holonomic (aka. P-finite, P-recursive or D-recursive) \(\textiff\) there exists~a holonomic recurrence for it, i.e.

@@

Note that the sum can only take~a finite number of subsequent elements of the sequence \(A\) into the relation.

@@

\begin{example}
    Sequences \(a_n = \sqrt{n}\), \(b_n = n^n\) are \textbf{not} holonomic, i.e.~none of them satisfies~a linear recurrence equation with polynomial coefficients~\cite{nonholo}.
\end{example}

\begin{definition}[\cite{complexity}]
    \label{def:holo_2}
    Equivalently to Definition~\ref{def:holo_1}, \(A = (a_i)_{i=0}^{\infty}\) is holonomic \(\textiff\) there exists~a generating function in the following form:
    \[\sum_{i=0}^{r} p_i(z) \gf{A}^{(i)}(z) = b(z)\]
    for some fixed \(r \in \N\); \(b, p_0, \ldots, p_r \in \poly\) not all identical to zero.
\end{definition}

The equivalence of these two definitions assures us that if we find~an generating function equation describing the desired objects satisfying Definition~\ref{def:holo_2}, we will be able to find holonomic recurrence mentioned in Definition~\ref{def:holo_1}~\cite{complexity}.

\begin{example}
    \label{ex:gfun-rec}
    For generating function counting full binary trees defined in Example~\ref{ex:bin_gf}, i.e.:
    \[\gf{B}(z) = 1 + z\gf{B}^2(z)\]
    we can use GFUN~\cite{gfun} to immediately find~a holonomic recurrence of our interest:

@@

Combinatorial interpretation is just~a fancy name for something you probably know and use. For example if you know the binomial coefficient \(\binom{n}{k}\), I bet it was introduced to you as something like: "let's assume you have \(n\) distinguishable balls. If you want to choose \(k\) from them, then you can do it in \(\binom{n}{k}\) ways" and now, whenever you see the term \(\binom{n}{k}\), you interpret it as the number of ways to choose \(k\) objects from the set of size \(n\). This is the combinatorial interpretation of the binomial coefficient.

Holonomic recurrences are very useful if we want to count objects of~a given size. Of course, size can be defined in more complex ways, e.g.~we may want to find the number of binary trees with \(n\) internal nodes having exactly \(m\) nodes being left child of its parent. Unfortunately, there is no commonly known method to find their intuitive interpretation in the world of the objects they describe.

In the thesis we try to gather methods and results that may be helpful for anyone trying to find such~an interpretation, as we believe that it may give better insight into the nature of the structure and make it easier to study the structure's properties.

@@

In the thesis, if not stated otherwise, we will be working on rooted ordered unlabeled \(n\)-ary trees. To understand the terms, let's define them:
\begin{itemize}
    \item rooted tree is~a tree with single vertex marked as root. This allows us to define hierarchy of nodes using such terms as parent and child;
    \item ordered trees introduce order on children from the leftmost to the rightmost;
    \item unlabeled trees have nodes that are not distinguishable by their label. We will introduce soon kinds of nodes, but you shouldn't treat them as labels, but rather way to describe the structure of the node that is already present (e.g.~we can call node with just left child~an \(l\) node, but it is not its label. If the right child was added to the \(l\) node, it would not be~an \(l\) node anymore).
\end{itemize}
Additionally we allow that \(n\)-ary node has some of its children unset (e.g.~we can define~a binary node with just left child set, but not right one).

In order to make notation easier and unambiguous, we introduce pre-order notation of~an \(n\)-ary tree.

@@

Here, \(\enc{v}\) is defined as follows:
\begin{itemize}
    \item if \(v\) is~an existing node, \(\enc{v} = \n\);
    \item if \(v\) is~a non-existing node (i.e.~we describe non-existing child of some node), \(\enc{v} = \no\).
\end{itemize}

@@

You can note that the encoding of~a given tree is unique. Unfortunately, not every sequence encodes proper tree.

We can also define prefixes and suffixes of \(T\) in relation to~a node \(v\).

\begin{definition}
    \(\tSub(v)\) is defined as~a whole subtree rooted with node \(v\).
\end{definition}

\begin{definition}
    Let \(v\) be~a node described using \(k\)-th symbol of \(\enc{T}\). Notation of the prefix of \(T\) in relation to \(v\), namely \(\enc{\tPre(T, v)}\), is defined as~a prefix of \enc{T} of size \(k - 1\).
\end{definition}

\begin{definition}
    Let \(n\) be the length of the sequence \(\enc{T}\). Notation of suffix of \(T\) in relation to \(v\), namely \(\enc{\tSuf(T, v)}\), is defined as~a suffix of \enc{T} of size \(n - \size{\enc{\tPre(T, v)}} - \size{\enc{\tSub(v)}}\).
\end{definition}

In other words, prefix of the tree \(T\) in relation to \(v\) describes~a part of the tree that appears before subtree rooted in \(v\) in pre-order notation. Suffix describes part of the tree that appears after it.

@@

Throughout thesis we define some operations using this notation. Valid pre-order notation of \(n\)-ary tree (note that not all sequences are valid pre-order notations) can be easily translated back into~a tree. In order to do so, we just take first symbol of the notation and create the root accordingly. Then for each of the \(n\) children of the root, in sequence, we decode it recursively as if it was whole tree. Each such operation consumes some prefix of the notation. At the end we should be left with no symbols in the sequence.

@@

We define the size of~a tree as the number of its internal nodes (i.e.~nodes that are not leaves). Our generating function is therefore~a solution of the following equation

@@

After~a quick transformation, we get the following holonomic recurrence:

@@

It is easy to see that~a tree of size \(n\) (i.e.~with \(n\) internal nodes) has \(n + 1\) leaves and therefore \(2n + 1\) nodes in total. Analogically the tree of size \(n + 1\) has \(n + 2\) leaves and \(2n + 3\) nodes in total. If you see some resemblance to our holonomic recurrence, you are perfectly right.

We are going to create~a bijection between the left and right side of the recurrence. We will be using the term \textit{bijective language} to talk about the recurrence using the objects it describes and their transformations.
\todo{rewrite}

\begin{interpretation}
Let's translate the recurrence to the bijective language.
\begin{itemize}
    \item \(b_n\) is simply the set of all full binary trees of size \(n\),
    \item \((n + 2) b_{n + 1}\) is the set of all full binary trees of size \(n + 1\) with one of the leaves pointed,
    \item \((2n + 1) b_{n}\) is the set of all full binary trees of size \(n\) with one of the nodes (either internal or~a leaf) pointed (i.e.~distinguished),
    \item \(2\) at the beginning of \(2 (2n + 1) b_{n}\) indicates that we will be using \(2\) copies of the set defined in the previous bullet.
\end{itemize}

@@

Let's define two local operations for tree \(T\) with~a single node \(v\) pointed: \(R_l(T, v)\) and \(R_r(T, v)\). The term \textit{local} means that the operation can operate on just some finite fragment of~a tree with predefined size (e.g.~node and its 2 children, but cannot traverse the whole tree).

@@

We can now take the bijective interpretation of \((2n + 1) B(n)\), i.e.~trees of size \(n\) with~a single node pointed. For each of such trees, \(T\) we create new one by applying \(R_l\) and \(R_r\) to the pair \(T', v\), where \(T'\) is \(T\) with removed pointers and \(v\) is~a node that is pointed in \(T\). Therefore we generate \(2(2n + 1) B(n)\) trees of size \(n + 1\).
\todo{rewrite}

@@

In the second case it is enough to notice that \(R_l\) generates trees with pointed nodes being~a left child, and \(R_r\) generates trees with pointed nodes being right child. That immediately provides us with the conclusion that we are generating no duplicates.

@@

\begin{definition}
    \label{def:unary_binary_tree}
    An unary-binary tree is~an ordered tree in which any node can have either 0, 1 or 2 children.
\end{definition}

We do not allow nodes with gaps here, e.g.~a binary node with just one child or unary node with no children.

We will define size as number of edges. Note that~a tree having \(n\) edges has exactly \(n + 1\) nodes.

@@

And with~a bit of Maple magic~\cite{gfun}, we immediately get the following recurrence:

@@

Definition~\ref{def:unary_binary_tree} is precise, but introduces~a new kind of node that we would have to take care of. We can avoid it by using~a trick to make unary-binary trees recall binary trees. We do not change the structure, just the notation.

\begin{definition}
    \label{def:unary_binary_2}
    A left-leaning binary tree is~a binary tree with the following constraint added: if~a node has~a right child, it also has~a left child. It can have just the left child.
\end{definition}

The bijection between unary-binary trees and left-leaning binary ones is very simple. We just represent unary nodes as nodes with just~a left child. The other cases translate immediately to themselves. Please note that this transformation preserves the size of the tree (i.e.~number of edges).

@@

\begin{definition}
    Virtual node is~an artificially defined node that does not affect the structure definition (i.e.~they are not included when counting the number of children of the node).
\end{definition}

\begin{definition}
    Actual node is~a node that is not virtual.
\end{definition}

\begin{definition}
    A virtual leaf is~a virtual node added to the node in place of~a non-existing child.
\end{definition}

@@

Please note that \(n + 2\) matches the multiplier of \(u_n\) in holonomic equation. The set \((n + 2) U(n)\) corresponds to trees of size \(n\) with single virtual leaf pointed.

\begin{definition}
    We say that node \(v\) is~an virtual-oneling \(\textiff\) \(v\) has no virtual siblings. Note that the root is oneling even if it has no parent.
\end{definition}

@@

\begin{definition}
    If \(u\) is parent of \(v\), we say that sibling of \(u\) is~an uncle of \(v\).
\end{definition}

@@

\begin{enumerate}
    \item Vro - virtual leaf being right virtual-oneling;
    \item Vrt - virtual leaf being right virtual-twinling. We will divide this case into:
        \begin{enumerate}
            \item VrtUrv - Vri which uncle is~a right virtual leaf;
            \item VrtUra - Vri which uncle is~a right actual node;
            \item VrtUla - Vri which uncle is~a left actual node.
        \end{enumerate}
    \item Vlt - virtual leaf left virtual-twinling. Note that from definition \ref{def:unary_binary_2}, left virtual leave cannot have~a sibling being~an actual node. We can divide also this case into:
        \begin{enumerate}
            \item VltUrv - Vlt which uncle is~a right virtual leaf;
            \item VltUra - Vlt which uncle is~a right actual node;
            \item VltUla - Vlt which uncle is~a left actual node;
        \end{enumerate}
    \item A - Actual node.
\end{enumerate}

@@

Let's start from \(U(n-1)\). Its multiplier is \(2n + 1\). It suggests that we can point to any node, as in~a tree having \(n - 1\) actual edges, we have \(n\) actual and \(n + 1\) virtual nodes.

In the case of \(U(n-2)\), the multiplier is \(3(n-1)\), which suggests that we should perform \(3\) distinct types of transformations on the trees with~a pointed actual node.

In the table below we show transformations that are applied to trees from \(U(n-1)\) and \(U(n-2)\) in order to achieve trees from \(U(n)\). We mark pointed node with \(\pointed{}\). You can find the case name in the buttom left corners of cells.

@@

\begin{definition}
    \label{def:schroder_tree}
    A Schröder tree is~a tree with no unary nodes. We do not allow empty children here (e.g.~binary node with just one child).
\end{definition}

@@

This definition is precise, but is difficult to implement as~a structural recursion. Let's define binary structure that will allow us to encode all Schröder trees.

\begin{definition}
    \label{def:well_weighted}
    A well-weighted full binary tree is~a weighted tree such that:
    \begin{itemize}
        \item leaves are unweighted;
        \item internal nodes can have weight either \(1\) or \(2\);
        \item if~an internal node has weight 2, then its right child is not~a leaf.
    \end{itemize}
\end{definition}

Let's represent node \(v\) with~a weight \(n\) as \(v\weighted{n}\).

@@

Let's define the size of the tree as~a number of its leaves. As leaves is translated to leaves and internal nodes to internal nodes, the bijection \(\Phi\) is size-preserving.

@@

\(\gf{T}_1\) and \(\gf{T}_1\) are generating functions of trees rooted in nodes of weight \(1\) and \(2\) respectively. \(\gf{T}\) is generating function for the whole well-weighted full binary tree.

To be able to use Maple to generate the corresponding recurrence, we need to transform this system of equations to~a single one. We can eliminate \(\gf{T}_1(z)\) first.

@@

> algeqtodiffeq(%, T(z));
                               2            /d      ~                   2
    {z - 1 + (-z + 3) T(z) + (z  - 6 z + 1) |-- T(z)|, T(0) = RootOf(2 _Z  - _Z)}
                                            \dz     /

@@

We will show the bijection between two sides of the recursion. We will transform each tree from \((2 n - 1) T(n)\) in three ways in order to achieve all trees from \((n + 1) T(n + 1)\). We will get some extra ones which are not well-weighted, but we will deal with them in~a minute.

Let's define three transformations, \(L_1\), \(L_2\), and \(R_1\) that transform tree with~a pointed node \(s\) into bigger tree, as follows. We use dashed edges 
\begin{minipage}{1.5em}
\includesvg[width=\textwidth]{intro__dashed_edge}
\end{minipage}
to depict edges that may or may not exist (i.e.~when the node can have parent or child, but it does not have to). Whenewer we use solid edge, we expect this edge to actually exist.

@@

It also allows us to define tree representation of the terms.

>>>>

We also try to find a~combinatorial interpretation of the holonomic recurrence for trees of lambda terms in the de Bruijn notation.

@@

Every recursive structure can be described using a~generating function, which can be then transformed into a~holonomic recurrence. It allows us to quickly count objects of given structure with given size. At first glance, the recurrence seems to be closely related to the structure itself. It just seems like it should describe simple transformations resulting in constructing bigger objects from slightly smaller ones. In case of trees it could be for example adding vertices. However, there is no known method of interpreting it in the language of the combinatorial structures.

@@

Throughout the thesis we try to accumulate known results and techniques that may help solving the problem of finding combinatorial interpretations. We also propose an~approach that helped us to find an~intermediate solution for the problem of finding combinatorial interpretation for holonomic recurrence of lambda trees in the unary de Bruijn notation. We use the de Bruijn indices in order to get rid of letter-based variable names. Unary notation allows us to connect size of the tree to the values of the de Bruijn indices.

@@

Finally, in Section~\ref{sec:further_work}, we present our view on what can be done with the problem in the future. We believe that one day the problem will be fully solved and a~more generic approach to interpreting holonomic recurrences will emerge.

@@

\(\poly\) is a~set of all polynomials.

@@

\(V(T)\) is implicitly defined as a~set of nodes of the tree \(T\).

\(E(T)\) is implicitly defined as a~set of edges of the tree \(T\).

@@

\(e\) is implicitly assumed to be an~edge.

\(\troot(T)\) is implicitly defined as a~root of tree \(T\).

\(\tleft(v)\) is implicitly defined as a~left child of node \(v\).

\(\tright(v)\) is implicitly defined as a~right child of node \(v\).

\(\tSub(v)\) is implicitly defined as a~whole subtree rooted with node \(v\).

\(\tLeft(v)\) is implicitly defined as a~left subtree of node \(v\).

\(\tRight(v)\) is implicitly defined as a~right subtree of node \(v\).

@@

Uncle of node \(v\) is defined as a~sibling of the parent of \(v\).

We say that node \(v\) is an~oneling \(\textiff\) \(v\) has no siblings. Note that the root is oneling even if it has no parent.

@@

Leaf is a~node which has no children.

\(\n\) is implicitly defined as a~default identifier of a~node.

\(\no\) is implicitly defined as an~identifier of a~non-existing node (e.g.~if \(v\) is a~leaf, \(\tleft(v) = \n\)).

@@

Virtual node is an~artificially defined node that does not affect the structure definition (e.g.~we can define virtual leafs that are children of actual leaves of the binary tree. Such virtual leaves are not included when counting the number of children of the node, so the actual leaves are still called leaves).

Actual node is a~node that is not virtual.

We say that node \(v\) is an~virtual-oneling \(\textiff\) \(v\) has no siblings being virtual nodes.

@@

For a~binary tree, we say that node \(v\) is right (left) \(\textiff\) \(v\) is a~right (left) child of its parent.

For a~binary tree, we say that node \(v\) has a~right (left) parent \(u\) \(\textiff\) \(v\) is a~left (right) child of \(u\).

\(\pointed{v}\) is implicitly defined as a~pointed node \(v\). We can treat it as a~node with an~additional label \(\pointed{}\).

\(v\weighted{n}\) is implicitly defined as a~weighted node \(v\) with weight \(n\).

\(x_i\) is implicitly assumed to be an~\(i\)-th element of the sequence \(X\), where \(x\) can be any lower-case letter and \(X\) is upper-case \(x\).

Calligraphic \(\gf{X}\) is implicitly assumed to be a~generating function of the sequence \(X\) where \(X\) is any upper-case letter.

@@

We will be analyzing structures that can be expressed using recursive formulas. For example we can define a~full binary tree recursively as a~node having no children (i.e.~being a~leaf) or having exactly two subtrees being full binary trees. To express the disjoint alternative in the construction rule, we use \(+\). The reason will become obvious soon.

@@

\begin{definition}
The ordinary generating function of sequence \(A = (a_i)_{i=0}^{\infty}\) is defined as a~formal power series:
\[\gf{A}(z) = \sum_{i=0}^{\infty} a_i z^i .\]
\end{definition}

Note that in order to compute the \(k\)-th element of the sequence, namely \(a_k\), we just need to find a~constant corresponding to indeterminate \(z^k\).

If we have a sequence counting some objects and a corresponding generating function, we say that this function counts the objects.

\begin{example}
    \label{ex:bin_gf}
    Let's consider recursive definition of binary trees shown in Figure~\ref{fig:binary_recursion}. We can create a~functional equation to count trees with exactly n internal nodes. To do so, we make internal nodes contribute to the generating function with \(z^1 = z\) and leaves with \(z^0 = 1\):

@@

In order to get some intuition on how it works underneath, let's take a~look at how we can find the number of full binary trees of size \(2\). We can expand the recurrence a~bit:

@@

To visualize even more, let's take a~look at the line marked with \((*)\). The highlighted terms describe following tree:

@@

\begin{definition}[\cite{holotoolkit}]
    \label{def:holo_1}
    An infinite sequence \(A = (a_i)_{i=0}^{\infty}\) is holonomic (aka. P-finite, P-recursive or D-recursive) \(\textiff\) there exists a~holonomic recurrence for it, i.e.

@@

Note that the sum can only take a~finite number of subsequent elements of the sequence \(A\) into the relation.

@@

\begin{example}
    Sequences \(a_n = \sqrt{n}\), \(b_n = n^n\) are \textbf{not} holonomic, i.e.~none of them satisfies a~linear recurrence equation with polynomial coefficients~\cite{nonholo}.
\end{example}

\begin{definition}[\cite{complexity}]
    \label{def:holo_2}
    Equivalently to Definition~\ref{def:holo_1}, \(A = (a_i)_{i=0}^{\infty}\) is holonomic \(\textiff\) there exists a~generating function in the following form:
    \[\sum_{i=0}^{r} p_i(z) \gf{A}^{(i)}(z) = b(z)\]
    for some fixed \(r \in \N\); \(b, p_0, \ldots, p_r \in \poly\) not all identical to zero.
\end{definition}

The equivalence of these two definitions assures us that if we find an~generating function equation describing the desired objects satisfying Definition~\ref{def:holo_2}, we will be able to find holonomic recurrence mentioned in Definition~\ref{def:holo_1}~\cite{complexity}.

\begin{example}
    \label{ex:gfun-rec}
    For generating function counting full binary trees defined in Example~\ref{ex:bin_gf}, i.e.:
    \[\gf{B}(z) = 1 + z\gf{B}^2(z)\]
    we can use GFUN~\cite{gfun} to immediately find a~holonomic recurrence of our interest:

@@

Combinatorial interpretation is just a~fancy name for something you probably know and use. For example if you know the binomial coefficient \(\binom{n}{k}\), I bet it was introduced to you as something like: "let's assume you have \(n\) distinguishable balls. If you want to choose \(k\) from them, then you can do it in \(\binom{n}{k}\) ways" and now, whenever you see the term \(\binom{n}{k}\), you interpret it as the number of ways to choose \(k\) objects from the set of size \(n\). This is the combinatorial interpretation of the binomial coefficient.

Holonomic recurrences are very useful if we want to count objects of a~given size. Of course, size can be defined in more complex ways, e.g.~we may want to find the number of binary trees with \(n\) internal nodes having exactly \(m\) nodes being left child of its parent. Unfortunately, there is no commonly known method to find their intuitive interpretation in the world of the objects they describe.

In the thesis we try to gather methods and results that may be helpful for anyone trying to find such an~interpretation, as we believe that it may give better insight into the nature of the structure and make it easier to study the structure's properties.

@@

In the thesis, if not stated otherwise, we will be working on rooted ordered unlabeled \(n\)-ary trees. To understand the terms, let's define them:
\begin{itemize}
    \item rooted tree is a~tree with single vertex marked as root. This allows us to define hierarchy of nodes using such terms as parent and child;
    \item ordered trees introduce order on children from the leftmost to the rightmost;
    \item unlabeled trees have nodes that are not distinguishable by their label. We will introduce soon kinds of nodes, but you shouldn't treat them as labels, but rather way to describe the structure of the node that is already present (e.g.~we can call node with just left child an~\(l\) node, but it is not its label. If the right child was added to the \(l\) node, it would not be an~\(l\) node anymore).
\end{itemize}
Additionally we allow that \(n\)-ary node has some of its children unset (e.g.~we can define a~binary node with just left child set, but not right one).

In order to make notation easier and unambiguous, we introduce pre-order notation of an~\(n\)-ary tree.

@@

Here, \(\enc{v}\) is defined as follows:
\begin{itemize}
    \item if \(v\) is an~existing node, \(\enc{v} = \n\);
    \item if \(v\) is a~non-existing node (i.e.~we describe non-existing child of some node), \(\enc{v} = \no\).
\end{itemize}

@@

You can note that the encoding of a~given tree is unique. Unfortunately, not every sequence encodes proper tree.

We can also define prefixes and suffixes of \(T\) in relation to a~node \(v\).

\begin{definition}
    \(\tSub(v)\) is defined as a~whole subtree rooted with node \(v\).
\end{definition}

\begin{definition}
    Let \(v\) be a~node described using \(k\)-th symbol of \(\enc{T}\). Notation of the prefix of \(T\) in relation to \(v\), namely \(\enc{\tPre(T, v)}\), is defined as a~prefix of \enc{T} of size \(k - 1\).
\end{definition}

\begin{definition}
    Let \(n\) be the length of the sequence \(\enc{T}\). Notation of suffix of \(T\) in relation to \(v\), namely \(\enc{\tSuf(T, v)}\), is defined as a~suffix of \enc{T} of size \(n - \size{\enc{\tPre(T, v)}} - \size{\enc{\tSub(v)}}\).
\end{definition}

In other words, prefix of the tree \(T\) in relation to \(v\) describes a~part of the tree that appears before subtree rooted in \(v\) in pre-order notation. Suffix describes part of the tree that appears after it.

@@

Throughout thesis we define some operations using this notation. Valid pre-order notation of \(n\)-ary tree (note that not all sequences are valid pre-order notations) can be easily translated back into a~tree. In order to do so, we just take first symbol of the notation and create the root accordingly. Then for each of the \(n\) children of the root, in sequence, we decode it recursively as if it was whole tree. Each such operation consumes some prefix of the notation. At the end we should be left with no symbols in the sequence.

@@

We define the size of a~tree as the number of its internal nodes (i.e.~nodes that are not leaves). Our generating function is therefore a~solution of the following equation

@@

After a~quick transformation, we get the following holonomic recurrence:

@@

It is easy to see that a~tree of size \(n\) (i.e.~with \(n\) internal nodes) has \(n + 1\) leaves and therefore \(2n + 1\) nodes in total. Analogically the tree of size \(n + 1\) has \(n + 2\) leaves and \(2n + 3\) nodes in total. If you see some resemblance to our holonomic recurrence, you are perfectly right.

We are going to create a~bijection between the left and right side of the recurrence. We will be using the term \textit{bijective language} to talk about the recurrence using the objects it describes and their transformations.
\todo{rewrite}

\begin{interpretation}
Let's translate the recurrence to the bijective language.
\begin{itemize}
    \item \(b_n\) is simply the set of all full binary trees of size \(n\),
    \item \((n + 2) b_{n + 1}\) is the set of all full binary trees of size \(n + 1\) with one of the leaves pointed,
    \item \((2n + 1) b_{n}\) is the set of all full binary trees of size \(n\) with one of the nodes (either internal or a~leaf) pointed (i.e.~distinguished),
    \item \(2\) at the beginning of \(2 (2n + 1) b_{n}\) indicates that we will be using \(2\) copies of the set defined in the previous bullet.
\end{itemize}

@@

Let's define two local operations for tree \(T\) with a~single node \(v\) pointed: \(R_l(T, v)\) and \(R_r(T, v)\). The term \textit{local} means that the operation can operate on just some finite fragment of a~tree with predefined size (e.g.~node and its 2 children, but cannot traverse the whole tree).

@@

We can now take the bijective interpretation of \((2n + 1) B(n)\), i.e.~trees of size \(n\) with a~single node pointed. For each of such trees, \(T\) we create new one by applying \(R_l\) and \(R_r\) to the pair \(T', v\), where \(T'\) is \(T\) with removed pointers and \(v\) is a~node that is pointed in \(T\). Therefore we generate \(2(2n + 1) B(n)\) trees of size \(n + 1\).
\todo{rewrite}

@@

In the second case it is enough to notice that \(R_l\) generates trees with pointed nodes being a~left child, and \(R_r\) generates trees with pointed nodes being right child. That immediately provides us with the conclusion that we are generating no duplicates.

@@

\begin{definition}
    \label{def:unary_binary_tree}
    An unary-binary tree is an~ordered tree in which any node can have either 0, 1 or 2 children.
\end{definition}

We do not allow nodes with gaps here, e.g. a~binary node with just one child or unary node with no children.

We will define size as number of edges. Note that a~tree having \(n\) edges has exactly \(n + 1\) nodes.

@@

And with a~bit of Maple magic~\cite{gfun}, we immediately get the following recurrence:

@@

Definition~\ref{def:unary_binary_tree} is precise, but introduces a~new kind of node that we would have to take care of. We can avoid it by using a~trick to make unary-binary trees recall binary trees. We do not change the structure, just the notation.

\begin{definition}
    \label{def:unary_binary_2}
    A left-leaning binary tree is a~binary tree with the following constraint added: if a~node has a~right child, it also has a~left child. Of course a node with just a left child can appear in such tree.
\end{definition}

The bijection between unary-binary trees and left-leaning binary ones is very simple. We just represent unary nodes as nodes with just a~left child. The other cases translate immediately to themselves. Please note that this transformation preserves the size of the tree (i.e.~number of edges).

@@

\begin{definition}
    Virtual node is an~artificially defined node that does not affect the structure definition (i.e.~they are not included when counting the number of children of the node).
\end{definition}

\begin{definition}
    Actual node is a~node that is not virtual.
\end{definition}

\begin{definition}
    A virtual leaf is a~virtual node added to the node in place of a~non-existing child.
\end{definition}

@@

Please note that \(n + 2\) matches the multiplier of \(u_n\) in holonomic equation.

We will be using \(U(n)\) to describe sets of trees corresponding to \(u_n\) from the holonomic recurrence. The set \((n + 2) U(n)\) can be interpreted as trees of size \(n\) with single virtual leaf pointed.

\begin{definition}
    We say that node \(v\) is an~virtual-oneling \(\textiff\) \(v\) has no virtual siblings. Note that the root is an oneling even if it has no parent.
\end{definition}

@@

\begin{definition}
    If \(u\) is parent of \(v\), we say that sibling of \(u\) is an~uncle of \(v\).
\end{definition}

@@

\begin{enumerate}
    \item Vro - virtual leaf being right virtual-oneling;
    \item Vrt - virtual leaf being right virtual-twinling. We will divide this case into:
        \begin{enumerate}
            \item VrtUrv - Vri which uncle is a~right virtual leaf;
            \item VrtUra - Vri which uncle is a~right actual node;
            \item VrtUla - Vri which uncle is a~left actual node.
        \end{enumerate}
    \item Vlt - virtual leaf left virtual-twinling. Note that from definition \ref{def:unary_binary_2}, left virtual leave cannot have a~sibling being an~actual node. We can divide also this case into:
        \begin{enumerate}
            \item VltUrv - Vlt which uncle is a~right virtual leaf;
            \item VltUra - Vlt which uncle is a~right actual node;
            \item VltUla - Vlt which uncle is a~left actual node;
        \end{enumerate}
    \item A - Actual node.
\end{enumerate}

@@

Let's start from \(U(n-1)\). Its multiplier is \(2n + 1\). It suggests that we can point to any node, as in a~tree having \(n - 1\) actual edges, we have \(n\) actual and \(n + 1\) virtual nodes.

In the case of \(U(n-2)\), the multiplier is \(3(n-1)\), which suggests that we should perform \(3\) distinct types of transformations on the trees with a~pointed actual node.

In the table below we show transformations that are applied to trees from \(U(n-1)\) and \(U(n-2)\) in order to achieve trees from \(U(n)\). We mark pointed node with \(\pointed{}\). You can find the case name in the bottom left corners of cells.

@@

\begin{definition}
    \label{def:schroder_tree}
    A Schröder tree is a~tree with no unary nodes. Empty children are not allowed (e.g.~binary node with just one child).
\end{definition}

@@

This definition is precise, but is difficult to implement as a~structural recursion. Let's define binary structure that will allow us to encode all Schröder trees.

\begin{definition}
    \label{def:well_weighted}
    A well-weighted full binary tree is a~weighted tree such that:
    \begin{itemize}
        \item leaves are unweighted;
        \item internal nodes can have weight either \(1\) or \(2\);
        \item if an~internal node has weight 2, then its right child is not a~leaf.
    \end{itemize}
\end{definition}

Let's represent node \(v\) with a~weight \(n\) as \(v\weighted{n}\).

@@

Let's define the size of the tree as a~number of its leaves. As leaves are translated to leaves and internal nodes to internal nodes, the bijection \(\Phi\) is size-preserving.

@@

\(\gf{T}_1\) and \(\gf{T}_1\) are generating functions of trees rooted in nodes of weight \(1\) and \(2\) respectively. \(\gf{T}\) is a generating function for the whole well-weighted full binary tree.

To be able to use Maple to generate the corresponding recurrence, we need to transform this system of equations to a~single one. We can eliminate \(\gf{T}_1(z)\) first.

@@

> algeqtodiffeq(%, T(z));
                               2            /d      \                   2
    {z - 1 + (-z + 3) T(z) + (z  - 6 z + 1) |-- T(z)|, T(0) = RootOf(2 _Z  - _Z)}
                                            \dz     /

@@

We will show the bijection between two sides of the recursion. We will transform each tree from \((2 n - 1) T(n)\) in three ways in order to achieve all trees from \((n + 1) T(n + 1)\). We will get some extra ones which are not well-weighted, but we will deal with them in a~minute.

Let's define three transformations, \(L_1\), \(L_2\), and \(R_1\) that transform tree with a~pointed node \(s\) into bigger tree, as follows. We use dashed edges 
\begin{minipage}{1.5em}
\includesvg[width=\textwidth]{intro__dashed_edge}
\end{minipage}
to depict edges that may or may not exist (i.e.~when the node can have parent or child, but it does not have to). Whenever we use solid edge, we expect this edge to actually exist.

@@

It also allows us to define tree representation of the terms from \(\LambdadB\).
\todo{introduce unary notation; correlation between n and the size}
